Wait-free Queues with Polygarithmic Step Complexity

### 2022-07-27 article/psuedocode.tex
review get method

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_{this,b,i}$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element} \label{getBaseCase}
\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_{this,b,i}$ is in the left child of this node.}
\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+left.blocks[blocks[b].end\sub{left}-1].sum\sub{enq}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
\State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i+right.blocks[blocks[b].end\sub{right}-1].sum\sub{enq}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
\State \Return\tt{right.}\Call{GetEnq}{subBlock, i} 
\EndIf
\EndFunction{GetEnq}

@@

\begin{lemma}[Get BSearch correctness] \label{getSearch}
Preconditions of invocation of \nf{BSearch} in \nf{GetENQ} are satisfied i.e. if the direct subblock containg $E_{n,b,i}$ is computed in both cases in Lines \ref{leftChildGet} and \ref{rightChildGet}.
\end{lemma}
\begin{proof}
There are two cases that the subblock we're looking for is in the left child or the right child. If $i$th \nf{enqueue} of \nf{n.blocks[b]} was in the left child it would be in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by definition \ref{def::subblock}. \nf{sum\sub{enq}} is the sum of the number of enqueues in the prefix of a block. So $E_{n,b,i}$ is $E_{n.left, i+n.left.blocks[n.blocks[b].end_{left}-1]}$ which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$. The parameters in \ref{leftChildGet} are correct so  this \nf{BSearch} returns the subblock containing $E_{n,b,i}$.

If the enqueue we were looking for was in the right child as there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.right.blocks[n.blocks[b].end\sub{right}-1]} (Line \ref{rightChildGet}). Other parameters are the same as the left child.
\end{proof}

\begin{lemma}[Get correctness] \label{get}
If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of the tree. The base case, the leaves of the tree, is pretty straight forward (Line \ref{getBaseCase}). Since leaf blocks contain exactly one operation thus only \nf{GetENQ(b,1)} can be called on a leaf block \nf{b}. \nf{l.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$. In the next paragraph we prove \nf{n.GetENQ()} returns $E_{n,b,i}$, if \nf{n.child.GetENQ()} works properly.

For non leaf nodes it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n} is in the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n} or in the \nf{n.blocks[b]}'s subblocks in the right child of \nf{n} (line \ref{leftOrRight}). From Definition \ref{ordering} we know operations in a block are ordered by their process id and since in leaves of the tree are ordered ascendingly by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}). Furthermore \nf{b.num\sub{enq-left}} stores the number of \nf{enqueue()} operations from the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n}. So if \nf{i} is greater than \nf{b.num\sub{enq-left}} it means $i$th operation is propagated from the right child, otherwise we should search for the $i$th enqueue in the left child. By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} which their range is \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. We can compute the direct subblock containing the enqueue we are finding for with binary search by Lemma \ref{getSearch}. Finaly \nf{n.child.GetENQ()} is invoked on the subblock containing $E_{n,b,i}$ which returns $E_{n,b,i}$ by the hypothesis of the induction.
\end{proof}

@@

  \caption{The number and ordering of the enqueues propagated from the left and the right child to \nf{n.blocks[b]}. \nf{n.blocks[b]} is  colored in the tope node by the color of the subblocks propagted to it. Left subblocks in red come before the right subblocks in blue and \nf{n.blocks[b].sum\sub{enq}=n.blocks[b].sum\sub{enq-left}+n.blocks[b].sum\sub{enq-right}}.
  }\label{figGet}
\end{figure}

>>>>

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_{this,b,i}$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element} \label{getBaseCase}
\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_{this,b,i}$ is in the left child of this node.}
\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
\State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i+right.blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
\State \Return\tt{right.}\Call{GetEnq}{subBlock, i} 
\EndIf
\EndFunction{GetEnq}

@@

\begin{lemma}[Get correctness] \label{get}
If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of node \nf{n}. The base case, where \nf{n} os a leaf, is straightforward (Line \ref{getBaseCase}). Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called where \nf{n} is a leaf. \nf{n.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf \nf{n}.

For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$, if \nf{n.child.GetENQ(b,i)} returns $E_{n.child,b,i}$. For non-leaf nodes it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n} is in the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n} or in the \nf{n.blocks[b]}'s subblocks in the right child of \nf{n} (line \ref{leftOrRight}). From Definition \ref{ordering} we know enqueue operations in a block are ordered by their process id and since in leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}). Furthermore \nf{b.num\sub{enq-left}} stores the number of \nf{enqueue()} operations from the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n}. So if \nf{i} is greater than \nf{b.num\sub{enq-left}} it means $i$th operation is propagated from the right child, otherwise we should search for the $i$th enqueue in the left child. By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.

If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}. Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_{n,b,i}$ is $E_{n.left, i+eb}$, which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ search for $i+eb$th enqueue in \tt{n.left} and $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in \ref{leftChildGet} are for searching $E_{n.left, i+eb}$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_{n,b,i}$.

Else if the enqueue we are looking for is in the right child then there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.blocks[b-1].sum\sub{enq-right}} (Line \ref{rightChildGet}). Other parameters are assigned similar for the left child. So in both cases the direct subblock containing $E_{n,b,i}$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.

 Finally, \nf{n.child.GetENQ()} is invoked on the subblock containing $E_{n,b,i}$ which returns $E_{n,b,i}$ by the hypothesis of the induction.
\end{proof}

@@

  \caption{The number and ordering of the enqueues propagated from the left and the right child to \nf{n.blocks[b]}. Enqueue operations from the left subblocks, colored red, are ordered before the enequeue operations from the right child, colored blue.
  }\label{figGet}
\end{figure}
